A SAT-Based Counterexample Guided Method for Unbounded Synthesis

### 2016-05-08 paper.tex
Leonid's edits

The reactive synthesis problem is EXPTIME-complete so
na\"ive algorithms are infeasible on anything but simple systems.
% In this way the environment is equivalent to an existential search of
%the game and the controller is universal. Much of the complexity of reactive
%synthesis stems from managing the interactions of the alternating quantifiers
%on the state space of the game.
There are several techniques that aim to mitigate this complexity by
representing states and transitions of the system symbolically~\cite{Piterman_PS_06,bloem2014,morgenstern2013}. 
These techniques incrementally construct a symbolic representation of the set of 
discovered winning states of the game.  The downside of this approach is that 
keeping track of all discovered winning states can lead to a space explosion even
when using efficient symbolic representation such as BDDs or CNFs.

@@

We evaluate our method on the benchmarks of the 2015 synthesis competition
(SYNTCOMP'15).  While our solver solves fewer total instances than competitors,
it solves the largest number of unique instances for a sequential solver, i.e.,
instances that could not be solved by any other solver.  These results confirm
that there exist classes of problems that are hard for traditional synthesis
techniques, but can be efficiently solved by our method.  While further
performance improvements are clearly needed, in its current state the method
may be a worthwhile addition to a portfolio solver.
%%%While further performance improvements are clearly needed, in its current state the method 
%%%can be used in a portfolio solver to increase the number of tractable instances.

@@

A \emph{safety game}, $G = \langle X, U, C, \delta, I, E \rangle$, is defined
over boolean state variables $X$, uncontrollable action variables $U$, and
controllable action variables $C$.  $I$ is the initial state of the game given
as a valuation of state variables.  $E(X)$ is the set of error states
represented by its characteristic formula.  The transition relation $\delta(X,
U, C, X')$ of the game is a boolean formula that relates current state and
action to the set of possible next states of the game.  We assume deterministic
games, where $\delta(x,u,c,x'_1) \land \delta(x,u,c,x'_2) \implies
(x'_1=x'_2)$.

@@

A \emph{controller strategy} $\pi^c : (2^X, 2^U) \to 2^C$ is a mapping of states and
uncontrollable inputs to controllable actions. A controller strategy is winning
in a bounded game of maximum bound $\kappa$ if all runs $(x_0, u_0, \pi^c(x_0,
u_0)), (x_1, u_1, \pi^c(x_1, u_1)) \dots$ $(x_n, u_n, \pi^c(x_n, u_n))$ are
winning.  Bounded \emph{realisability} is the problem of determining the
existence of such a strategy for a bounded game.

An \emph{environment strategy} $\pi^e : 2^X \to 2^U$ is a mapping of states to
uncontrollable actions. A bounded run is winning for the environment if $x_0 =
I \land \exists i \in \{1..n\} (E(x_i))$ and an environment strategy is winning
for a bounded game if all runs $(x_0, \pi^e(x_0), c_0), (x_1, \pi^e(x_1), c_1)
\dots (x_n, \pi^e(x_n), c_n)$ are winning for the environment.  Safety games
are zero sum, therefore the existence of a winning controller strategy implies
the nonexistence of a winning environment strategy and vice versa.

@@

We introduce a running example to assist the explanation. We consider a simple
arbiter system in which the environment makes a request for a number of
resources (1 or 2), and the controller may grant access to up to two resources.
The total number of requests grows each round by the number of environment
requests and shrinks by the number of resources granted by the controller in
the previous round.  The controller must ensure that the number of unhandled
requests does not accumulate to more than 2.  Figure~\ref{fig:example} shows
the variables (\ref{fig:examplevars}) and the formulas for computing next-state
variables (\ref{fig:exampletrans}) for this example. We use primed identifiers
to denote next-state variables and curly braces to define the domain of a
variable.

This example is the $n=2$ instance of the more general problem of an arbiter of
$n$ resources. The set of winning states has no compact representation, which
makes the problem hard for BDD solvers. In Section~3 we will outline how the
unbounded game can be solved without enumerating all winning states.

@@

The bounded synthesis algorithm consists of two competing solvers, one for the
controller and one for the environment, that solve abstractions of the safety
game. Each solver builds a candidate strategy for its corresponding player
consisting of assignments to $2^C$ and $2^U$ respectively.

An abstraction of the game restricts actions available to one of the players.
Specifically, we consider abstractions represented as trees of actions,
referred to as \emph{abstract game trees} (AGTs).  Figure~\ref{fig:agt} shows
an example abstract game tree restricting the environment (abstract game trees
restricting the controller are similar).  In the abstract game, the controller
can freely choose actions whilst the environment is required to pick actions
from the tree.  After reaching a leaf, the enivironment continues playing
unrestricted.  The tree in Figure~\ref{fig:agt} restricts the first environment
action to \texttt{request=1}. At the leaf of the tree the game continues
unrestricted.

@@

Given an environment (controller) abstract game tree $T$ a \emph{partial
strategy} $Strat: \textsc{nodes}(T) \rightarrow C$ ($Strat: \textsc{nodes}(T)
\rightarrow U$) labels each node of the tree with the controller's
(environment's) action to be played in that node.   Given a partial strategy
$Strat$, we can map each leaf $l$ of the abstract game tree to $\langle
s',i'\rangle=\textsc{outcome}(\langle s, i\rangle, Strat, l)$ obtained by
playing all controllable and uncontrollable actions on the path from the root
to the leaf.

@@

    %% Make it clear that this is intuition only

    To solve our example, the algorithm begins by invoking a SAT solver to find
    a trace of length $k$ (here we consider $k=3$) from the initial state to an
    error state (see Figure~\ref{fig:trace}).  Given a trace we make an initial
    conjecture that the first action, \texttt{req=1}, is winning for the
    environment. This conjecture is captured in the abstract game tree shown in
    Figure~\ref{fig:agt}.  Next the solver attempts to find a counterexample to
    this conjecture by finding a trace that avoids the error states with the
    first environment action fixed to \texttt{req=1}, i.e. by playing the
    abstract game on behalf of the controller.

    The resulting trace is shown in Figure~\ref{fig:trace2}. The environment's
    candidate is refined using this trace and Figure~\ref{fig:agt2} shows the
    refined game tree.  The controller's counterexample was to grant two
    resources in the first round, to which the environment responds by now
    making two requests in the second round.

    Eventually, the refinements to the environment's candidate will force the
    controller to grant more resources and result in a game tree such as in
    Figure~\ref{fig:refined}.

@@

@@

The algorithm is organised as a counterexample-guided abstraction refinement
(CEGAR) loop.  The first step of the algorithm uses the \textsc{findCandidate}
function, described below, to come up with a candidate partial strategy that is
winning when the opponent is restricted to $T$.  If it fails to find a
strategy, this means that no winning strategy exists against the opponent
playing its partial strategy represented by $T$.  If, on the other hand, a
candidate partial strategy is found, we need to verify if it is indeed winning
for the abstract game $T$.

The \textsc{verify} procedure searches for a \emph{spoiling} counterexample
strategy in each leaf of the candidate partial strategy by calling
\textsc{solveAbstract} for the opponent. The dual solver solves games on behalf
of the controller player.

@@

It is possible to set a maximum bound such that all runs in the unbounded game will be
considered. The na\"ive approach is to use size of the state space as the bound
($2^{|X|}$) so that all states may be explored by the algorithm. A more nuanced
approach is to use the diameter of the game \cite{biere1999}, which is the
smallest number $d$ such that for any state $x$ there is a path of length $\leq
d$ to all other reachable states. However, the diameter is difficult to
estimate and can still lead to infeasibly long games.

@@

    One option is using a BDD solver, working backwards from the error set, to
    find all losing states. One iteration of this operation on our example
    would give the set: $\texttt{nrequests} = 3 \lor (\texttt{nrequests} = 2
    \land ( \texttt{resource0} = 0 \lor \texttt{resource1} = 0)) \lor
    (\texttt{nrequests} = 1 \land (\texttt{resource0} = 0 \land
    \texttt{resource1} = 0))$.  In the general case there is no compact
    representation of the losing set, so we try to avoid computing it by
    employing interpolation instead. The potential benefit of interpolation is
    that an interpolant can approximate the losing states in such a way that
    there is just enough information to prove realisability. It is also cheap
    to compute an interpolant using a SAT solver.
\end{exmpInt}

>>>>

The reactive synthesis problem is EXPTIME-complete (when starting 
from a symbolic representation of the game graph) so
na\"ive algorithms are infeasible on anything but simple systems.
% In this way the environment is equivalent to an existential search of
%the game and the controller is universal. Much of the complexity of reactive
%synthesis stems from managing the interactions of the alternating quantifiers
%on the state space of the game.
There are several techniques that aim to mitigate this complexity by
representing states and transitions of the system symbolically~\cite{Piterman_PS_06,bloem2014,morgenstern2013}. 
These techniques incrementally construct a symbolic representation of the set of 
discovered winning states of the game.  The downside of this approach is that 
keeping track of all discovered winning states can lead to a space explosion even
when using efficient symbolic representation such as BDDs or CNFs.

@@

We evaluate our method on the benchmarks of the 2015 synthesis competition
(SYNTCOMP'15).  While our solver solves fewer total instances than competitors,
it solves the largest number of unique instances, i.e.,
instances that could not be solved by any other sequential solver.  
These results confirm
that there exist classes of problems that are hard for traditional synthesis
techniques, but can be efficiently solved by our method.  While further
performance improvements are clearly needed, in its current state the method
may be a worthwhile addition to a portfolio solver.
%%%While further performance improvements are clearly needed, in its current state the method 
%%%can be used in a portfolio solver to increase the number of tractable instances.

@@

A \emph{safety game}, $G = \langle X, U, C, \delta, I, E \rangle$, is defined
over boolean state variables $X$, uncontrollable action variables $U$, and
controllable action variables $C$.  
We use $\mathcal{X}$, $\mathcal{U}$, and $\mathcal{C}$ to denote sets of valuations of 
variables $X$, $U$, and $C$ respectively.
$I$ is the initial state of the game given
as a valuation of state variables.  $E(X)$ is the set of error states
represented by its characteristic formula.  The transition relation $\delta(X,
U, C, X')$ of the game is a boolean formula that relates current state and
action to the set of possible next states of the game.  We assume deterministic
games, where $\delta(x,u,c,x'_1) \land \delta(x,u,c,x'_2) \implies
(x'_1=x'_2)$.

@@

A \emph{controller strategy} $\pi^c : \mathcal{X} \times \mathcal{U} \to \mathcal{C}$ 
is a mapping of states and
uncontrollable inputs to controllable actions. A controller strategy is winning
in a bounded game of maximum bound $\kappa$ if all runs $(x_0, u_0, \pi^c(x_0,
u_0)), (x_1, u_1, \pi^c(x_1, u_1)) \dots$ $(x_n, u_n, \pi^c(x_n, u_n))$ are
winning.  Bounded \emph{realisability} is the problem of determining the
existence of such a strategy for a bounded game.

An \emph{environment strategy} $\pi^e : \mathcal{X} \to \mathcal{U}$ is a mapping of states to
uncontrollable actions. A bounded run is winning for the environment if $x_0 =
I \land \exists i \in \{1..n\} (E(x_i))$ and an environment strategy is winning
for a bounded game if all runs $(x_0, \pi^e(x_0), c_0), (x_1, \pi^e(x_1), c_1)
\dots (x_n, \pi^e(x_n), c_n)$ are winning for the environment.  Safety games
are zero sum, therefore the existence of a winning controller strategy implies
the nonexistence of a winning environment strategy and vice versa.

@@

We introduce a running example to assist the explanation. We consider a simple
arbiter system in which the environment makes a request for a number of
resources (1 or 2), and the controller may grant access to up to two resources.
The total number of requests grows each round by the number of environment
requests and shrinks by the number of resources granted by the controller in
the previous round.  The controller must ensure that the number of unhandled
requests does not accumulate to more than 2.  Figure~\ref{fig:example} shows
the variables (\ref{fig:examplevars}), the initial state of the system (\ref{fig:exampleinit}), 
and the formulas for computing next-state
variable assignments (\ref{fig:exampletrans}) for this example. We use primed identifiers
to denote next-state variables and curly braces to define the domain of a
variable.

This example is the $n=2$ instance of the more general problem of an arbiter of
$n$ resources. For large values of $n$, the set of winning states has no compact representation, which
makes the problem hard for BDD solvers. In Section~3 we will outline how the
unbounded game can be solved without enumerating all winning states.

@@

Our bounded synthesis algorithm constructs abstractions of the game 
that restrict actions available to one of the players.
Specifically, we consider abstractions represented as trees of actions,
referred to as \emph{abstract game trees} (AGTs).  Figure~\ref{fig:agt} shows
an example abstract game tree restricting the environment (abstract game trees
restricting the controller are similar).  In the abstract game, the controller
can freely choose actions whilst the environment is required to pick actions
from the tree.  After reaching a leaf, the environment continues playing
unrestricted.  The tree in Figure~\ref{fig:agt} restricts the first environment
action to \texttt{request=1}. At the leaf of the tree the game continues
unrestricted.

@@

Given an environment (controller) abstract game tree $T$ a \emph{partial
strategy} $Strat: \textsc{nodes}(T) \rightarrow \mathcal{C}$ ($Strat: \textsc{nodes}(T)
\rightarrow \mathcal{U}$) labels each node of the tree with the controller's
(environment's) action to be played in that node.   Given a partial strategy
$Strat$, we can map each leaf $l$ of the abstract game tree to $\langle
s',i'\rangle=\textsc{outcome}(\langle s, i\rangle, Strat, l)$ obtained by
playing all controllable and uncontrollable actions on the path from the root
to the leaf.  An environment (controller) partial strategy is \emph{winning against $T$} 
if all its outcomes are states that are winning for the environment (controller)
in the concrete game.

@@

    %% Make it clear that this is intuition only
    We present the intuition behind our bounded synthesis method by applying
    its \emph{simplified version} to the running example.
    We begin by finding 
    a trace of length $k$ (here we consider $k=3$) that is winning for the environment, 
    i.e., that starts from the initial state and reaches an
    error state (see Figure~\ref{fig:trace}).  We use a SAT solver to find such a trace,
    precisely as one would do in bounded model checking.
    Given this trace we make an initial
    conjecture that any trace starting with action \texttt{req=1} is winning for the
    environment. This conjecture is captured in the abstract game tree shown in
    Figure~\ref{fig:agt}.  We validate this conjecture by searching for a counterexample
    trace that avoids the error state with the
    first environment action fixed to \texttt{req=1}.   
    Such a trace, that refutes the conjecture, is shown in Figure~\ref{fig:trace2}. 
    In this trace, the controller wins by responding with \texttt{gr0=1, gr1=1} in the first round.
    This move represents controller's partial strategy against the 
    abstract game tree in Figure~\ref{fig:agt}.  This partial strategy is shown in Figure~\ref{}.

    Next we strengthen the abstract game tree taking this partial strategy into account.
    To this end we again use a SAT solver to find a trace where the environment wins while
    the controller plays according to the partial strategy.  In the resulting trace (now shown), 
    the environment plays \texttt{req=2} in the second round.  We refine the abstract game tree
    using this move as shown in Figure~\ref{fig:agt2}.  The controller's partial strategy was to grant two
    resources in the first round, to which the environment responds by now
    making two requests in the second round.

    When the environment cannot refine the tree by extending existing branches, it backtracks 
    and creates new branches.
    Eventually, we obtain the abstract game tree shown in Figure~\ref{fig:refined} for which 
    there does not exist a winning partial strategy on behalf of the controller.  We conclude that 
    the game is winning for the environment.

%    Eventually, the refinements to the environment's candidate will force the
%    controller to grant more resources and result in a game tree such as in
%    Figure~\ref{fig:refined}.

@@

The full bounded synthesis algorithm is more complicated: upon finding a candidate 
partial strategy on behalf of player $p$ against abstract game tree $T$, it first checks whether the strategy is winning 
against $T$.  By only considering such strong candidates, we reduce the number of 
refinements needed to solve the game.  To this end, the algorithm checks whether each 
outcome of the candidate strategy is a winning state for $\textsc{opponent}(p)$ by recursively 
invoking the synthesis algorithm on behalf of the opponent.  Thus, our bounded
synthesis algorithm can be seen as running two competing solvers, for the
controller and for the environment.

%The solvers build candidate partial strategy 
%for their corresponding players, which are used to refine the abstractions.

@@

The algorithm is organised as a counterexample-guided abstraction refinement
(CEGAR) loop.  The first step of the algorithm uses the \textsc{findCandidate}
function, described below, to come up with a candidate partial strategy that is
winning when the opponent is restricted to $T$.  If it fails to find a
strategy, this means that no winning partial strategy exists against the opponent
playing according to $T$.  If, on the other hand, a
candidate partial strategy is found, we need to verify if it is indeed winning
for the abstract game $T$.

The \textsc{verify} procedure searches for a \emph{spoiling} counterexample
strategy in each leaf of the candidate partial strategy by calling
\textsc{solveAbstract} for the opponent. The dual solver solves games on behalf
of the opponent player.

@@

It is possible to set a maximum bound such that all runs in the unbounded game will be
considered. The na\"ive approach is to use size of the state space as the bound
(${|\mathcal{X}|}$) so that all states may be explored by the algorithm. A more nuanced
approach is to use the diameter of the game \cite{biere1999}, which is the
smallest number $d$ such that for any state $x$ there is a path of length $\leq
d$ to all other reachable states. However, the diameter is difficult to
estimate and can still lead to infeasibly long games.

@@

    One option is using a BDD solver, working backwards from the error set, to
    find all losing states. One iteration of this operation on our example
    would give the set: $\texttt{nrequests} = 3 \lor (\texttt{nrequests} = 2
    \land ( \texttt{resource0} = 0 \lor \texttt{resource1} = 0)) \lor
    (\texttt{nrequests} = 1 \land (\texttt{resource0} = 0 \land
    \texttt{resource1} = 0))$.  In the general case there is no compact
    representation of the losing set, so we try to avoid computing it by
    employing interpolation instead. The benefit of interpolation is
    that it allows approximating the losing states efficiently by obtaining 
    an interpolant from a SAT solver.
\end{exmpInt}
